---
title: "02_get_commute_distances"
author: "Addison Larson"
date: "Started 20221119 // Last knitted `r format(Sys.time(), '%Y%m%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: yeti
knit: (function(inputFile, encoding) {
      rmarkdown::render(inputFile,
                        encoding = encoding,
                        output_dir = "C:/Users/ALarson/My_Projects/chi_town/Programs",
                        output_file = "02_get_commute_distances.html") })
---

**Purpose:** Compute distance and time traveling from Chicago census tract centroids to Mathematica office using Google Maps Distance Matrix API. 

# Setup

```{r setup}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(here)
library(sf)
library(ggmap)
library(lubridate)

root <- here()
raw  <- file.path(root, "Data/Raw")

projection <- "+proj=tmerc +lat_0=36.66666666666666 +lon_0=-88.33333333333333 +k=0.9999749999999999 +x_0=300000.0000000001 +y_0=0 +ellps=GRS80 +datum=NAD83 +to_meter=0.3048006096012192 +no_defs"

register_google(scan(file.path(root, "Keys/google_maps_api_key.txt"), what = "character"))

```

# Compute and set params

- Time of day (set to depart at 10:00 a.m. Chicago time the Wednesday after whenever this program is run)
- Office address
- Compute centroids of Cook County census tracts

```{r set_params}

# Set to correct time zone so times are correct
Sys.setenv(TZ = "America/Chicago")

# Identify "next Wednesday" based on current time
this_weekday_text <- wday(ymd(today()), label = TRUE)
print(this_weekday_text)
this_weekday_num <- wday(ymd(today()))

if (this_weekday_num > 4) {
  add_days <- 11 - this_weekday_num
} else if (this_weekday_num < 4) {
  add_days <- 4 - this_weekday_num
} else {
  add_days <- 7
}

next_wednesday <- ymd(today()) + ddays(add_days)
print(ymd(next_wednesday))

inject_string <- paste0("departure_time=", as.numeric(as.POSIXct(paste0(next_wednesday, " 8:00"))))



# Office address
ofc <- "111 E Wacker Dr, Chicago, IL 60601"



# Compute tract centroids
tract_centroids <- st_read(file.path(raw, "tract_boundaries.gpkg")) %>%
  st_transform(projection) %>%
  st_centroid(.) %>%
  st_transform(4326) %>%
  mutate(lon = st_coordinates(.)[,"X"],
         lat = st_coordinates(.)[,"Y"],
         coords = paste0(lat, ", ", lon)) %>%
  st_set_geometry(NULL)

```

# Compute distances and export

Make sure to append census tract GEOID before export.

```{r compute_distances}

walk_data <- mapdist(
  from = tract_centroids$coords,
  to = ofc,
  mode = "walking",
  output = "all",
  inject = inject_string
)

# Replace names of list elements from centroids with GEOID
# This was initially written with the assumption that mapdist return order matches row order of input data (tract_centroids) -- this appears to be wrong
list_names <- data.frame(
  names = names(walk_data),
  order = 1:length(walk_data)
) %>%
  left_join(., tract_centroids, by = c("names" = "coords")) %>%
  select(names, order, GEOID) %>%
  arrange(order)

# Note that first two printouts align but third does not
print(head(list_names))
print(head(names(walk_data)))
print(head(tract_centroids$GEOID))

names(walk_data) <- list_names$GEOID

saveRDS(walk_data, file = file.path(raw, "commute_dist_walk.Rds"))



transit_data <- mapdist(
  from = tract_centroids$coords,
  to = ofc,
  mode = "transit",
  output = "all",
  inject = inject_string
)

# Replace names of list elements from centroids with GEOID
# This was initially written with the assumption that mapdist return order matches row order of input data (tract_centroids) -- this appears to be wrong
list_names <- data.frame(
  names = names(transit_data),
  order = 1:length(transit_data)
) %>%
  left_join(., tract_centroids, by = c("names" = "coords")) %>%
  select(names, order, GEOID) %>%
  arrange(order)

# Note that first two printouts align but third does not
print(head(list_names))
print(head(names(transit_data)))
print(head(tract_centroids$GEOID))

names(transit_data) <- list_names$GEOID

saveRDS(transit_data, file = file.path(raw, "commute_dist_transit.Rds"))

```
