---
title: "10_clean_rawdata"
author: "Addison Larson"
date: "Started 20221114 // Last knitted `r format(Sys.time(), '%Y%m%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: yeti
knit: (function(inputFile, encoding) {
      rmarkdown::render(inputFile,
                        encoding = encoding,
                        output_dir = "C:/Users/ALarson/My_Projects/chi_town/Programs",
                        output_file = "10_clean_rawdata.html") })
---

**Purpose:** Clean up the data files downloaded in `01_download_rawdata.R` and saved in `./Data/Raw` so they're easier to combine into a tract-level analysis file later. 

# Setup

```{r setup}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(sf)
library(assertr)
library(here)
library(ggplot2)
library(stringr)
library(janitor)

root  <- here()
raw   <- file.path(root, "Data/Raw")
input <- file.path(root, "Data/Input")

projection <- "+proj=tmerc +lat_0=36.66666666666666 +lon_0=-88.33333333333333 +k=0.9999749999999999 +x_0=300000.0000000001 +y_0=0 +ellps=GRS80 +datum=NAD83 +to_meter=0.3048006096012192 +no_defs"

```

# Clean data

## Neighborhood boundaries

```{r nhoods_boundaries}

nhoods_boundaries_raw <- st_read(file.path(raw, "nhoods_boundaries.geojson"))

glimpse(nhoods_boundaries_raw)

nhoods_boundaries <- nhoods_boundaries_raw %>%
  assert(is_uniq, pri_neigh) %>%
  select(pri_neigh) %>%
  rename(neighborhood = pri_neigh) %>%
  st_transform(projection)

city_boundaries <- nhoods_boundaries %>%
  summarize(geometry = st_union(geometry))

```

## Census tract boundaries

- Only retain census tracts intersecting with city boundary.
- Save a list of these GEOIDs to restrict median rent file.

```{r tract_boundaries}

tract_boundaries_raw <- st_read(file.path(raw, "tract_boundaries.gpkg"))

glimpse(tract_boundaries_raw)

tract_boundaries <- tract_boundaries_raw %>%
  assert(is_uniq, GEOID) %>%
  rename(geoid = GEOID) %>%
  st_transform(projection)

# Run intersection; convert results to vector with values 0 (Not in Chicago); 1 (In Chicago)
intersects_city <- st_intersects(tract_boundaries, city_boundaries) %>%
  lapply(., function(i) ifelse(length(i) == 0, 0, 1)) %>%
  unlist(.)

stopifnot(length(intersects_city) == nrow(tract_boundaries))

# Append to data frame
tract_boundaries$in_chicago <- intersects_city

table(tract_boundaries$in_chicago)

ggplot(data = tract_boundaries) +
  geom_sf(aes(fill = in_chicago))

tract_boundaries <- tract_boundaries %>%
  filter(in_chicago == 1) %>%
  select(geoid)

chicago_tracts <- tract_boundaries$geoid

```

## Median rent of census tracts

```{r median_rent}

median_rent_raw <- read.csv(file.path(raw, "median_rent.csv"))

glimpse(median_rent_raw)

median_rent <- median_rent_raw %>%
  assert(is_uniq, GEOID) %>%
  rename(
    geoid = GEOID,
    median_rent_est = estimate,
    median_rent_moe = moe
  ) %>%
  filter(geoid %in% chicago_tracts)

stopifnot(all(sort(median_rent$geoid) == sort(tract_boundaries$geoid)))

```

## Census tract centroids

These points may come in handy when computing distances.

```{r tract_centroids}

tract_centroids <- st_centroid(tract_boundaries)

plot(st_geometry(tract_boundaries))
plot(st_geometry(tract_centroids), add = TRUE, col = "blue", pch = 20)

```

## Parks locations

```{r parks}

parks_locations_raw <- st_read(file.path(raw, "parks_locations.geojson"))

glimpse(parks_locations_raw)

# Drop "unimproved" and "pending"
table(parks_locations_raw$park_class)

parks_locations <- parks_locations_raw %>%
  assert(is_uniq, park) %>%
  filter(!(grepl("^UNIMPROVED|^PENDING", park_class))) %>%
  select(park, label, park_class) %>%
  st_transform(., projection)

plot(st_geometry(city_boundaries))
plot(st_geometry(parks_locations), add = TRUE, col = "darkgreen")

```

## Library locations

```{r libraries}

library_locations_raw <- read.csv(file.path(raw, "library_locations.csv"))

# Need to parse location X, Y and convert file to spatial
glimpse(library_locations_raw)

library_locations <- library_locations_raw %>%
  assert(is_uniq, NAME) %>%
  # Verify pattern is "(num.num, -num.num)"
  verify(str_detect(LOCATION, "^\\(\\d+\\.\\d+, -\\d+\\.\\d+\\)")) %>%
  # Drop unneeded parens, then extract x and y from location column
  mutate(
    location_noparens = str_replace_all(LOCATION, "\\(|\\)", ""),
    y = str_split_fixed(location_noparens, ", ", 2)[,1],
    x = str_split_fixed(location_noparens, ", ", 2)[,2],
    y2 = y,
    x2 = x
  ) %>%
  st_as_sf(., coords = c("x2", "y2"), crs = 4326) %>%
  st_transform(., projection) %>%
  rename(name = NAME) %>%
  select(name, x, y)

plot(st_geometry(city_boundaries))
plot(st_geometry(library_locations), add = TRUE)

```

## Grocery locations

```{r grocery}

grocery_locations_raw <- read.csv(file.path(raw, "grocery_locations.csv")) %>%
  clean_names(.)

glimpse(grocery_locations_raw)

grocery_locations <- grocery_locations_raw %>%
  mutate(latitude2 = latitude,
         longitude2 = longitude) %>%
  st_as_sf(., coords = c("longitude2", "latitude2"), crs = 4326) %>%
  select(store_name, license_id, account_number, latitude, longitude) %>%
  st_transform(., projection)

plot(st_geometry(city_boundaries))
plot(st_geometry(grocery_locations), add = TRUE)

```

# Export

```{r export}

export_spatial <- function(obj) {
  
  obj_name <- paste0(obj, ".gpkg")
  
  get(obj) %>%
    st_transform(., 4326) %>%
    st_write(., file.path(input, obj_name), delete_dsn = TRUE)
  
}

lapply(
  c("nhoods_boundaries",
    "city_boundaries",
    "tract_boundaries",
    "tract_centroids",
    "parks_locations",
    "library_locations",
    "grocery_locations"),
  export_spatial
)

write.csv(median_rent, file.path(input, "median_rent.csv"), row.names = FALSE)

```
