---
title: "20_create_analysisfile"
author: "Addison Larson"
date: "Started 20221114 // Last knitted `r format(Sys.time(), '%Y%m%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: yeti
knit: (function(inputFile, encoding) {
      rmarkdown::render(inputFile,
                        encoding = encoding,
                        output_dir = "C:/Users/ALarson/My_Projects/chi_town/Programs",
                        output_file = "20_create_analysisfile.html") })
---

**Purpose:** Combine files in `./Data/Input` into a usable tract-level analysis file. At the time of writing this file will NOT contain all variables of interest and is a work in progress. 

# Setup

```{r setup}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(sf)
library(assertr)
library(here)
library(ggplot2)
library(stringr)
library(tools)
library(units)
library(matrixStats)
library(gdalUtilities)

root     <- here()
input    <- file.path(root, "Data/Input")
analysis <- file.path(root, "Data/Analysis") 

projection <- "+proj=tmerc +lat_0=36.66666666666666 +lon_0=-88.33333333333333 +k=0.9999749999999999 +x_0=300000.0000000001 +y_0=0 +ellps=GRS80 +datum=NAD83 +to_meter=0.3048006096012192 +no_defs"

# This function fixes parcel land use shapefile, otherwise I get error "Unknown WKB type 12"
# Barely adapted from StackOverflow: https://gis.stackexchange.com/a/389854
ensure_multipolygons <- function(shp) {
  
  # Make tempfiles to temporarily save shps to disk (ogr2ogr needs them)
  tmp1 <- tempfile(fileext = ".gpkg")
  tmp2 <- tempfile(fileext = ".gpkg")
  
  # Save inshp to disk
  st_write(shp, tmp1)
  
  # Convert to multipolygon
  ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
  
  shp_fixed <- st_read(tmp2)
  st_sf(st_drop_geometry(shp), geom = st_geometry(shp_fixed))
  
}

```

# Read in data files

```{r readin}

file_types <- file_ext(list.files(input))
stopifnot(all(file_types %in% c("csv", "gpkg")))

read_inputs <- function(obj) {
  
  # Determine file extension
  file_type <- file_ext(obj)
  if (!file_type %in% c("csv", "gpkg")) {
    stop(paste("Function not written to handle file type", file_type))
  }
  
  if (file_type == "gpkg") {
    st_read(file.path(input, obj)) %>%
      st_transform(., projection)
  } else {
    read.csv(file.path(input, obj))
  }
}

for (obj in list.files(input)) {
  
  # Determine file extension
  file_type <- file_ext(obj)
  
  # Drop file extension off of name
  obj_environment_name <- str_replace(obj, paste0("\\.", file_type, "$"), "")
  
  fl <- read_inputs(obj)
  assign(obj_environment_name, fl)
  
}

```

# Create potentially useful constructs

## Assign neighborhoods to tracts

```{r neighborhoods_merge}

overlay_tracts_nhoods <- st_intersection(tract_boundaries, nhoods_boundaries) %>%
  mutate(aland = st_area(.))

# Sometimes tracts are split over multiple neighborhoods. Assign based on largest overlap of land area.
# Show example, GEOID 17031020400
ex_nhoods <- filter(nhoods_boundaries, neighborhood %in% c("Rogers Park", "West Ridge"))
ex_tract  <- filter(overlay_tracts_nhoods, geoid == "17031020400")
# Land area in Rogers Park is infinitesimal
overlay_tracts_nhoods %>%
  st_set_geometry(NULL) %>%
  filter(geoid == "17031020400") %>%
  print(.)

ggplot() +
  geom_sf(data = ex_nhoods, fill = "gray50", alpha = 0.5) +
  geom_sf(data = ex_tract, aes(fill = neighborhood), alpha = 0.5)

dedup_tracts_nhoods <- overlay_tracts_nhoods %>%
  st_set_geometry(NULL) %>%
  group_by(geoid) %>%
  arrange(desc(aland)) %>%
  slice(., 1) %>%
  ungroup(.) %>%
  select(-aland)

# Show example after deduping
dedup_tracts_nhoods %>%
  filter(geoid == "17031020400") %>%
  print(.)

```

## Create parkland measurements

- Distance to nearest
- Area within a mile of tract centroid

```{r parks_measurements}

# Distance to nearest park ----------------------------------------------------

# Distance matrix is computed in feet (map units) by default; convert to miles after computing
parks_dist <- st_distance(tract_centroids, parks_locations) %>%
  set_units(mile) %>%
  drop_units(.)

# N rows of distance matrix = N tracts
stopifnot(nrow(parks_dist) == nrow(tract_centroids))

# Take row minimum to get distance to nearest
parks_mindist <- rowMins(parks_dist)

# Top-code outliers. Coding to 1.001 because we compute land area within 1-mile buffer below.
summary(parks_mindist)
plot(density(parks_mindist))

parks_mindist_topcode <- ifelse(parks_mindist > 1, 1.001, parks_mindist)

summary(parks_mindist_topcode)
plot(density(parks_mindist_topcode))

print(length(which(parks_mindist != parks_mindist_topcode)))



# Park area within 1 mi of tract centroid -------------------------------------

centroid_buffers <- st_buffer(tract_centroids, 5280)

plot(st_geometry(centroid_buffers))
plot(st_geometry(tract_centroids), pch = 16, add = TRUE)
plot(st_geometry(parks_locations), col = "darkgreen", add = TRUE)

overlay_parks_tracts <- st_intersection(parks_locations, centroid_buffers) %>%
  mutate(aland = st_area(.))

# Show example park and overlay
example_park <- filter(parks_locations, label == "Archer")
example_intersects <- filter(overlay_parks_tracts, label == "Archer")
example_tracts <- filter(centroid_buffers, geoid %in% unique(example_intersects$geoid))

ggplot() +
  geom_sf(data = example_park, fill = "darkgreen", alpha = 1) +
  geom_sf(data = example_tracts, fill = "blue", alpha = 0.1)

# Group by tract, compute total land area
dedup_parks_tracts <- overlay_parks_tracts %>%
  st_set_geometry(NULL) %>%
  group_by(geoid) %>%
  summarize(total_parkland = sum(aland)) %>%
  ungroup(.) %>%
  drop_units(.)

# No total land area should exceed total area of a circle with a 1-mile radius (buffer size)
stopifnot(all(dedup_parks_tracts$total_parkland < pi * (5280 ^ 2)))



# Combine data ----------------------------------------------------------------

parks_stats <- data.frame(
  geoid = tract_centroids$geoid,
  parks_mindist_topcode = parks_mindist_topcode
) %>%
  left_join(., dedup_parks_tracts, by = "geoid") %>%
  mutate(parks_aland_1mi = ifelse(is.na(total_parkland), 0, total_parkland / (5280 ^ 2))) %>% # Convert sq ft to sq mi
  select(-total_parkland)

# Verify that these two stats are consistent -- If tract wasn't within a mile of nearest park, aland within 1 mi should be 0!
stopifnot(all((parks_stats$parks_mindist_topcode == 1.001) == (parks_stats$parks_aland_1mi == 0)))

```

## Create library measurements

- Distance to nearest

```{r library_measurements}

# Distance matrix is computed in feet (map units) by default; convert to miles after computing
library_dist <- st_distance(tract_centroids, library_locations) %>%
  set_units(mile) %>%
  drop_units(.)

# N rows of distance matrix = N tracts
stopifnot(nrow(library_dist) == nrow(tract_centroids))

# Take row minimum to get distance to nearest
library_mindist <- rowMins(library_dist)

# Top-code outliers to 95th percentile.
summary(library_mindist)
plot(density(library_mindist))

library_mindist_topcode <- ifelse(library_mindist > quantile(library_mindist, 0.95), quantile(library_mindist, 0.95), library_mindist)

summary(library_mindist_topcode)
plot(density(library_mindist_topcode))

print(length(which(library_mindist != library_mindist_topcode)))

library_stats <- data.frame(
  geoid = tract_centroids$geoid,
  library_mindist_topcode = library_mindist_topcode
)

```

## Create grocery measurements

- Distance to nearest
- Count within a half-mile of tract centroid

```{r grocery_measurements}

# Distance to nearest grocery store -------------------------------------------

# Distance matrix is computed in feet (map units) by default; convert to miles after computing
grocery_dist <- st_distance(tract_centroids, grocery_locations) %>%
  set_units(mile) %>%
  drop_units(.)

# N rows of distance matrix = N tracts
stopifnot(nrow(grocery_dist) == nrow(tract_centroids))

# Take row minimum to get distance to nearest
grocery_mindist <- rowMins(grocery_dist)

# Top-code outliers to 95th percentile. This could (will?) lead to inconsistencies when computing area below.
summary(grocery_mindist)
plot(density(grocery_mindist))

grocery_mindist_topcode <- ifelse(grocery_mindist > quantile(grocery_mindist, 0.95), quantile(grocery_mindist, 0.95), grocery_mindist)

summary(grocery_mindist_topcode)
plot(density(grocery_mindist_topcode))

print(length(which(grocery_mindist != grocery_mindist_topcode)))



# Count of grocery stores within 0.5 mi of tract centroid ---------------------

centroid_buffers <- st_buffer(tract_centroids, 2640) # 2640 is 5280 ft (1 mi) / 2 -> 0.5 mi

plot(st_geometry(centroid_buffers))
plot(st_geometry(grocery_locations), col = "blue", pch = 16, add = TRUE)

overlay_grocery_tracts <- st_intersection(grocery_locations, centroid_buffers)

# Show example overlay
example_tract <- filter(tract_boundaries, geoid == "17031832100")
example_buffer <- filter(centroid_buffers, geoid == "17031832100")
example_intersects <- filter(overlay_grocery_tracts, geoid == "17031832100")

ggplot() +
  geom_sf(data = example_buffer, fill = "gray50", alpha = 0.5) +
  geom_sf(data = example_tract, fill = "blue", alpha = 1) +
  geom_sf(data = example_intersects, color = "black")

# Group by tract, compute count within each buffer
dedup_grocery_tracts <- overlay_grocery_tracts %>%
  st_set_geometry(NULL) %>%
  group_by(geoid) %>%
  summarize(n_grocery_0.5mi = n()) %>%
  ungroup(.)

print(filter(dedup_grocery_tracts, geoid == "17031832100"))



# Combine data ----------------------------------------------------------------

grocery_stats <- data.frame(
  geoid = tract_centroids$geoid,
  grocery_mindist_topcode = grocery_mindist_topcode
) %>%
  left_join(., dedup_grocery_tracts, by = "geoid") %>%
  mutate(n_grocery_0.5mi = ifelse(is.na(n_grocery_0.5mi), 0, n_grocery_0.5mi))

grocery_stats %>%
  filter(n_grocery_0.5mi == 0) %>%
  summary(grocery_mindist_topcode)

```

# Commute data

```{r commute_data}

walk_stats <- filter(commute_dist_walk, geoid %in% tract_boundaries$geoid)
stopifnot(all(sort(walk_stats$geoid) == sort(tract_boundaries$geoid)))
names(walk_stats)[!names(walk_stats) == "geoid"] <- paste0("walk_", names(walk_stats)[!names(walk_stats) == "geoid"])

transit_stats <- filter(commute_dist_transit, geoid %in% tract_boundaries$geoid)
stopifnot(all(sort(transit_stats$geoid) == sort(tract_boundaries$geoid)))
names(transit_stats)[!names(transit_stats) == "geoid"] <- paste0("transit_", names(transit_stats)[!names(transit_stats) == "geoid"])

```


# Residential mix

The download of the 2015 Land Use Inventory (https://www.cmap.illinois.gov/data/land-use/inventory) includes a PDF with land use classification codes

- Single-family are codes 1111 and 1112
- Multi-family is code 1130
- Urban mix with residential component is code 1216. ("Similar to 1215 Urban Mix, but in areas where there is a likelihood of one or more residential units included in the parcel on the upper floors of the building. In newer “mixed-use” developments, this category is applied if the residential and commercial components share a single parcel. In older urban areas characterized by a block of storefronts, all parcels in the block receive this code provided the majority of buildings are 2-4 stories high and that there is evidence (such as Census housing unit counts) that some of the buildings contain dwelling units in the upper floors. Buildings taller than 4 stories that are predominantly multi-family should be coded 1130 Multi-Family.")

```{r land_use_residential}

table(chicago_land_use_2015$LANDUSE)
stopifnot(all(!is.na(chicago_land_use_2015$LANDUSE)))

chicago_land_use_2015 <- chicago_land_use_2015 %>%
  mutate(
    land_use_residential = case_when(
      LANDUSE %in% c(1111, 1112) ~ "single_family",
      LANDUSE %in% c(1116, 1130) ~ "multi_family",
      TRUE ~ "drop"
    )
  )

land_use_singlefamily <- chicago_land_use_2015 %>%
  filter(land_use_residential == "single_family") %>%
  select(land_use_residential) %>%
  ensure_multipolygons(.) %>%
  st_union(.) %>%
  st_transform(., projection) %>%
  st_as_sf(., "MULTIPOLYGON")

land_use_multifamily <- chicago_land_use_2015 %>%
  filter(land_use_residential == "multi_family") %>%
  select(land_use_residential) %>%
  ensure_multipolygons(.) %>%
  st_union(.) %>%
  st_transform(., projection) %>%
  st_as_sf(., "MULTIPOLYGON")

overlay_singlefamily <- st_intersection(tract_boundaries, land_use_singlefamily)
overlay_multifamily  <- st_intersection(tract_boundaries, land_use_multifamily)

# Verify that geoid uniquely identifies overlay file.
stopifnot(length(unique(overlay_singlefamily$geoid)) == nrow(overlay_singlefamily))
stopifnot(length(unique(overlay_multifamily$geoid))  == nrow(overlay_multifamily))

overlay_singlefamily_flat <- overlay_singlefamily %>%
  mutate(singlefamily_aland = st_area(.)) %>%
  drop_units(.) %>%
  select(geoid, singlefamily_aland) %>%
  st_set_geometry(NULL)

overlay_multifamily_flat <- overlay_multifamily %>%
  mutate(multifamily_aland = st_area(.)) %>%
  drop_units(.) %>%
  select(geoid, multifamily_aland) %>%
  st_set_geometry(NULL)

tract_boundaries_aland <- tract_boundaries %>%
  mutate(tract_aland_for_landuse = st_area(.)) %>%
  drop_units(.) %>%
  select(geoid, tract_aland_for_landuse) %>%
  st_set_geometry(NULL)

# Merge on to tract shapefile to find overall percentage of land for the land use
tract_residential <- left_join(tract_boundaries_aland, overlay_singlefamily_flat, by = "geoid") %>%
  left_join(., overlay_multifamily_flat, by = "geoid") %>%
  mutate(
    singlefamily_aland = ifelse(is.na(singlefamily_aland), 0, singlefamily_aland),
    tract_pct_singlefamily = singlefamily_aland / tract_aland_for_landuse,
    multifamily_aland = ifelse(is.na(multifamily_aland), 0, multifamily_aland),
    tract_pct_multifamily = multifamily_aland / tract_aland_for_landuse
  )

```

# Combine and export

```{r combine_export}

merge_fn <- function(combined_df, add_df) {
  add_df %>%
    mutate(geoid = as.character(geoid)) %>%
    left_join(combined_df, .)
}

combined_data <- Reduce(
  merge_fn,
  list(
    tract_boundaries,
    dedup_tracts_nhoods,
    parks_stats,
    library_stats,
    grocery_stats,
    walk_stats,
    transit_stats,
    median_rent,
    tract_residential
  ),    
  accumulate = FALSE) %>%
  st_transform(., 4326)

str(combined_data)

ggplot(combined_data) +
  geom_sf(aes(fill = neighborhood)) +
  theme(legend.position = "none")

st_write(combined_data, file.path(analysis, "analysisfile.gpkg"), delete_dsn = TRUE)

```
