---
title: "30_suitability_analysis"
author: "Addison Larson"
date: "Started 20221120 // Last knitted `r format(Sys.time(), '%Y%m%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: yeti
knit: (function(inputFile, encoding) {
      rmarkdown::render(inputFile,
                        encoding = encoding,
                        output_dir = "C:/Users/ALarson/My_Projects/chi_town/Programs",
                        output_file = "30_suitability_analysis.html") })
---

**Purpose:** Use the analysis file to identify neighborhoods that are decently close to Mathematica office, don't cost \$\$\$ in rent, and ideally have other perks like a grocery store within walking distance. (This is a first draft; I ought to find better data that indicates the stock and price of studio and 1-bd units.)

# Setup

```{r setup}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(sf)
library(here)
library(ggplot2)
library(tibble)

root     <- here()
analysis <- file.path(root, "Data/Analysis") 

projection <- "+proj=tmerc +lat_0=36.66666666666666 +lon_0=-88.33333333333333 +k=0.9999749999999999 +x_0=300000.0000000001 +y_0=0 +ellps=GRS80 +datum=NAD83 +to_meter=0.3048006096012192 +no_defs"

data <- st_read(file.path(analysis, "analysisfile.gpkg")) %>%
  st_transform(projection)

```

# Preview layers

```{r show_layers}

layers <- setdiff(names(data), c("geoid", "geom"))

plot_fn <- function(lyr) {
  
  if (is.character(data[[lyr]]) & length(unique(data[[lyr]])) > 10) {
    suppress_legend <- TRUE
  } else {
    suppress_legend <- FALSE
  }
  
  ggplot() +
    geom_sf(data = data, aes(fill = get(lyr))) +
    labs(title = lyr) +
    theme(legend.title = element_blank()) +
    if (suppress_legend) theme(legend.position = "none")

}

lapply(layers, plot_fn)

```

# Assign (arbitrary, subject to change) evaluation vars / weights

Consider refactoring this so variables and weights aren't hard-coded in program.

```{r assign_vars_weights}

vars_weights <- tribble(
  ~var, ~wt,
  "median_rent_bool",      0.3,
  "walk_distance_bool",    0.2,
  "transit_duration_bool", 0.2,
  "parks_aland_1mi_scale", 0.1,
  "library_mindist_scale", 0.1,
  "n_grocery_0.5mi_scale", 0.1
)

stopifnot(sum(vars_weights$wt) == 1)

```

# Data prep

## Construct Boolean variables `walk_distance_bool`, `transit_duration_bool`, `median_rent_bool` (ie., "+ 1 if 2 mile or shorter walk from office")

```{r construct_booleans}

summary(data$walk_distance)
data$walk_distance_bool <- ifelse(data$walk_distance <= 2, 1, 0)
table(data$walk_distance_bool, useNA = "always")

summary(data$transit_duration)
data$transit_duration_bool <- ifelse(data$transit_duration <= 45, 1, 0)
table(data$transit_duration_bool, useNA = "always")

summary(data$median_rent_est)
data$median_rent_bool <- ifelse(data$median_rent_est <= 1250, 1, 0)
table(data$median_rent_bool, useNA = "always")

```

## Rescale

Literally dividing by max observed so that these variables range 0-1.

```{r rescale}

data <- data %>%
  mutate(
    across(c("parks_aland_1mi", "library_mindist_topcode", "n_grocery_0.5mi"),
           function(col) col / max(col, na.rm = TRUE),
           .names = "{.col}_scale")) %>%
  rename(library_mindist_scale = library_mindist_topcode_scale)


```

# Apply weights

This code is going to be extra ugly. Refactor somehow?

```{r apply_wts}

for (var_n in 1:nrow(vars_weights)) {
  
  var <- vars_weights$var[var_n]
  wt  <- vars_weights$wt[var_n]
  
  data$var_score <- data[[var]] * wt
  names(data)[which(names(data) == "var_score")] <- paste0(var, "_score")
  
}

data <- data %>%
  rowwise(.) %>%
  mutate(overall_score = sum(c_across(ends_with("score")))) %>%
  ungroup(.)

```

# Show results

```{r show_score}

layers <- c(paste0(vars_weights$var, "_score"), "overall_score")

lapply(layers, plot_fn)


```

# Aggregate and summarize by neighborhood

Let's just show the top ten.

```{r aggregate}

score_by_nhood <- data %>%
  st_set_geometry(NULL) %>%
  group_by(neighborhood) %>%
  summarize(avg_nhood_score = mean(overall_score, na.rm = TRUE)) %>%
  ungroup(.) %>%
  arrange(desc(avg_nhood_score)) %>%
  mutate(avg_nhood_rank = row_number())

print(as.data.frame(score_by_nhood))

data <- left_join(data, score_by_nhood, by = "neighborhood")

data %>%
  filter(avg_nhood_rank <= 10)


ggplot() +
  geom_sf(data = data, fill = "gray50") +
  geom_sf(data = filter(data, avg_nhood_rank <= 10),
          aes(fill = avg_nhood_rank))
  
```

# Export

```{r save}

st_write(data, file.path(analysis, "analysisfile_score.gpkg"), delete_dsn = TRUE)
write.csv(st_set_geometry(data, NULL), file.path(analysis, "analysisfile_score.csv"), row.names = FALSE)

```
